---
title: "BIOS0003 Field Work (conducted on 2026-10-07)"
output: 
  html_notebook:
    fig_caption: yes
    math_method:
      engine: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
editor_options: 
  chunk_output_type: inline
---

## 1 Code setup

*I use something like this in all Rmd files*

```{r}
library(tidyverse)
library(ggplot2)
library(tinytable)

# Set global figure dimensions (e.g., 8 inches wide by 5 inches high)
knitr::opts_chunk$set(
  fig.width = 4, 
  fig.height = 4,
  fig.align = "center" # Automatically centers all plots in the preview
)

options(tinytable_html_mathjax = TRUE)
theme_set(theme_minimal(base_size = 14, base_family = "Helvetica"))
```

--- 

## 2 Data Cleaning and Prep

To clean and prepare the data I took the following steps: 

1. I took the list from the list Maddy created in this [google doc](https://docs.google.com/document/d/1ayEZ5W-id2mDLf8K16_WwtTJwzxPaptZ_ai-xaE0v5s/edit?tab=t.0) and added it to this [google sheet](https://docs.google.com/spreadsheets/d/1wY-feIpz5OgWtJsYgT1v5Mk01S2WGyqa1xK2fIkk68g/edit?gid=0#gid=0).
1. I then munged the data so that there was complete data for the following columns: 
    1. Habitat
    1. Habitat grid number (taken from the diagram we have)
    1. Square type (either Small or Large)
    1. Species (the observed Species)
1. I then worked to identify species that had the same name but spelt differently and identified the following: 
    1. Grass and Grasses - both changed to Grass
    1. Hook lobed dandelions and Hooked lobed dandelions - both changed to Hook lobed dandelions
    1. Wild cherry and Wild cherry tree - both changed to Wild cherry
1. Assigned an Overall Grid Number (from 1 to 8) to each grid

The output is stored here

*./data/in/field_work_data_cleaned.csv*

Now we can load the data into a dataframe

```{r}
df = read_csv("../data/in/field_work_data_cleaned.csv", show_col_types = FALSE)
df
```

---

## 3 Overall Species Accumulation Curve 

>First, lets do this manually

we are going to need a data frame to hold the species accumulation data:

```{r}
df_accum <- tibble(
  survey_unit = c(0:8), # start at 0 and 8 sample units, overall_grid_number
  species_accum = rep(0, 9) # set species accumulation to 0 to start
)
df_accum
```

We also need a list to hold the species we have seen in total, well work through the sample units adding new species to this list and updating the species_accum in the dataframe above.

```{r}
species_all = c()
```

Lets do the first sample unit (overall_grid_number = 1) manually...

```{r}
# Pull a distinct list of species seen in ogn 1
species_ogn <- df |> 
  filter(overall_grid_number == 1) |> 
  distinct(Species) |> 
  pull(Species)

# Add only new ones to the species_all vector
species_all = union(species_all, species_ogn)

# Count how many species we have all together 
species_count = length(species_all)

# Update the df_accum data table for ogn = 1
# Note: we're using 1 + 1 here as we started at 0
df_accum$species_accum[1 + 1] <- species_count

df_accum
```

Now we have done that once, lets loop through the rest of the sample units and fill up the data frame:

```{r}
# Loop through the remaing ogn's from 2 to 8
for (ogn in 2:8) {
  # Then just do the same as we did for ogn = 1
  species_ogn <- df |> 
    filter(overall_grid_number == ogn) |> 
    distinct(Species) |> 
    pull(Species)

  # Add only new ones to the species_all vector
  species_all = union(species_all, species_ogn)
  
  # Count how many species we have all together 
  species_count = length(species_all)
  
  # Update the df_accum data table for ogn = 1
  df_accum$species_accum[ogn + 1] <- species_count
}

df_accum
```

And now let's plot this curve:

```{r}
df_accum |> 
  ggplot(mapping = aes(x = survey_unit, y = species_accum)) +
    geom_point() +
    stat_smooth(method = "loess", formula = "y ~ x") +
    labs(
      title = "Species Accumulation Curve",
      x = "Survey Unit (Overall Grid Number)",
      y = "Accumulated Species Count"
    )
```

**!!Boom, we have our first species accumulation curve!!**




